---
title: "Functional Flow Analysis for the San Joaquin River"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Incorperating Effective Functional Flows into R2R

The Reorienting to Recovery Structured Decision Making group proposed a effective functional flow (EFF) scenario. To date, there is no available CalSIM EFF run so we cannot model full EFF on the entire system. This document outlines our approach to piloting an EFF on the San Joaquin River constructed to maximize salmonid survival.

This analysis is based on the methods for the Sacramento River EFF (`vignettes/EFF.Rmd`) and we do not reiterate the methods in detail here except to explain where they deviate for the San Joaquin.

### Review of Sacramento River EFF methods

A synthetic EFF hydrograph was developed for the Sacramento River using the following high-level steps: 

* Using the [Natural Flows Database](https://ceff.ucdavis.edu/sites/g/files/dgvnsk5566/files/media/documents/CEFF%20Technical%20Report%20Ver%201.0%20Mar_31_2021_DRAFT_FINAL%20for%20web.pdf), [California Environmental Flows Framework](https://ceff.ucdavis.edu/sites/g/files/dgvnsk5566/files/media/documents/CEFF%20Technical%20Report%20Ver%201.0%20Mar_31_2021_DRAFT_FINAL%20for%20web.pdf), and [Modeling Functional Flows in California's Rivers](https://www.frontiersin.org/articles/10.3389/fenvs.2022.787473/full#:~:text=Functional%20flow%20components%20in%20California,ecological%20processes%20in%20riverine%20ecosystems) as a basis, a monthly hydrograph was created to mimic unimpaired historical flows

* Following this initial exercise, our synthetic hydrograph was modified to prioritize maximizing salmonid survival by accounting for flow-to-survival thresholds as outlined in [Michel 2021](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.3498).

* Because this initial approach did not maximize *habitat* for salmon, the hydrograph was further modified to a) maximize spawning habitat from Oct-Dec as defined [here](https://reorienting-to-recovery.github.io/DSMhabitat/reference/sacramento_habitat.html), b) set a base flow for dry season (Aug-Sep) juvenile survival informed by the Natural Flows database, c) optimize flows during the last three months of outmigration (May-July) using [Michel 2021](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.3498), and d) utilize additional flows for maximizing floodplain habitat from Jan-April.

### Adapting the Sacramento River EFF for the San Joaquin River

Our goals for developing an EFF hydrograph for the San Joaquin River were to maximize habitat and survival and reduce minimum flows when salmonids are not in the system.

In order to adapt the Sacramento EFF methodology, we need a flow-survival relationship for salmonids like those outlined in [Michel 2021](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.3498). Steps include: 

- scaling the existing flow-survival thresholds calculated by [Michel 2021](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.3498) for the Sacramento the San Joaquin
- modify the SJ survival model logic to be flow-based instead of temperature-based (so that the EFF has an effect)

### Methods

#### Scale flow-to-survival relationships

We compared the flows for the `Upper-mid Sacramento River` and `San Joaquin River` produced by the 2019 BiOp and calculated a monthly ratio of San Joaquin flows to Upper-mid Sacrmaento flows (`San Joaquin River flows / Upper-mid Sacramento River flows`). We then averaged that monthly proportion over the years modeled in the DSM (`1980:2000`). We used this proportion to scale the flow values required for survival rates of `3%`, `18.9%`, and `50.8%` for the San Joaquin.

```{r, echo = FALSE}
DSMflow::flows_cfs$biop_itp_2018_2019 |> 
  select(date, `Upper-mid Sacramento River`, `San Joaquin River`) |> 
  mutate(flow_relationship = `San Joaquin River` / `Upper-mid Sacramento River`) |> pivot_longer(`Upper-mid Sacramento River`:`San Joaquin River`,
               names_to = "watershed", values_to = "flow_cfs") |> 
  ggplot(aes(x = date, y = flow_cfs, color = watershed)) + 
  geom_line() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(y = "Flow (cfs)", 
       x = "",
       title = "Calsim BiOp 2018-2019 flows")

scale_thresholds <- DSMflow::flows_cfs$biop_itp_2018_2019 |> 
  select(date, `Upper-mid Sacramento River`, `San Joaquin River`) |> 
  mutate(flow_relationship = `San Joaquin River` / `Upper-mid Sacramento River`) |> pivot_longer(`Upper-mid Sacramento River`:`San Joaquin River`,
               names_to = "watershed", values_to = "flow_cfs") |> 
  filter(watershed == "San Joaquin River")

scale_thresholds |> 
  ggplot(aes(x = date, y = flow_relationship)) + 
  geom_line() +
  # geom_vline(aes(xintercept = c(1980)), linetype = "dashed", color = "blue") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(y = "Ratio", 
       x = "",
       title = "San Joaquin :Sacramento flow ratio")
# ranges from 0:1, 3 in some later years and model years

# filter to model years and then average to get scaling factor
san_joaquin_scaling_factor <- scale_thresholds |> 
  filter(year(date) %in% 1980:2000,
         month(date) %in% 5:7) |> 
  summarise(scaling_factor = mean(flow_relationship)) |> 
  pull(scaling_factor)

# now calculate flow required for survival using the scaling factor to
# adjust flows
flow_to_survival_thresholds <- tibble("Survival" = c(3, 18.9, 50.8),
                                      "Sacramento Flow Threshold (cfs)" = c(4279, 10712, 22872)) |> 
  mutate(`San Joaquin Flow Threshold` = round(`Sacramento Flow Threshold (cfs)` * san_joaquin_scaling_factor, 0))

knitr::kable(flow_to_survival_thresholds)
```

```{r, echo=FALSE}
# Helper / Set Up Code
# Compare total flow volumne function 
compare_total_flow_volumne <- function(calsim_flows, natural_flow_flows, proposed_eff){
  all_flows <- bind_rows(calsim_flows |> mutate(type = "calsim"), 
                         natural_flow_flows |> mutate(type = "Natural flows Database flows"),
                         proposed_eff |> mutate(type = "Proposed Model EFF")) |> 
    select(month, flow_cfs, type) |> 
    mutate(acre_ft_per_month = flow_cfs * 60.3707) |> #TODO confirm conversion 
    group_by(type) |> 
    summarise(annual_acre_feet = sum(acre_ft_per_month)) |> 
    rename(Scenario = type, "Annual Acre Ft Water" = annual_acre_feet) 
  
  return(knitr::kable(all_flows))
}

# Find Representative water year types
water_year_types <- waterYearType::water_year_indices |> 
  filter(location == "San Joaquin Valley", WY %in% c(1922:2003)) |> 
  select(year = WY, index = Index, year_type = Yr_type) 

median_wet_year_index <- water_year_types |> 
  filter(year_type %in% c("Wet", "Above Normal")) |> 
  pull(index) |> median()

represenative_wet_years <- water_year_types |>
  mutate(distance_from_median = abs(index - median_wet_year_index)) |> 
  filter(distance_from_median < 1) |> 
  arrange(distance_from_median)
# Representative Wet years are 1974 and 1937 - both .0650 away

median_dry_year_index <- water_year_types |> 
  filter(year_type %in% c("Critical", "Dry", "Below Normal")) |> 
  pull(index) |> median()

representative_dry_years <- water_year_types |>
  mutate(distance_from_median = abs(index - median_dry_year_index)) |> 
  filter(distance_from_median < 1) |> 
  arrange(distance_from_median)
# Representative Wet years are 1939 & 1959 - both .005 away

```

### Proposed Model EFF {.tabset}

#### Wet Year EFF

The table below shows monthly flow EFF for a wet year model EFF on the San Joaquin River. This table was constructed using the Natural Flows database and model survival thresholds.

```{r, echo = FALSE}
# WET YEAR EFF 
# for Natural Flows Tool, chose farthest downstream (can modify)
# NATURAL FLOWS TOOL COMID: 2821742

# For calsim flow picked wet year, wettest year on record 
wet_year <- DSMflow::flows_cfs$biop_itp_2018_2019 |> 
  select(date, flow_cfs = `San Joaquin River`) |> 
  filter(year(date) == 1983) |> # most flow by year
  mutate(water_year = as.Date(ifelse(month(date) %in% 10:12, 
                             date,
                             as.Date(paste0(year(date) + 1, "-", month(date), "-", day(date))))),
         month = factor(month.abb[month(date)], levels = month.abb[c(10, 11, 12, 1:9)])) |> 
  mutate(water_year = ifelse(is.na(water_year), as.Date("1996-02-28"), water_year),
         water_year = as.Date(water_year)) # fix leap year

# wet_year |>
#   ggplot(aes(x = water_year, y = flow_cfs)) +
#   geom_line()
```

```{r,echo = FALSE}
# TODO left off here trying to figure out how Erin pulled from 
# the natural flows database to reconstruct this (checking against EFF.Rmd)

# right now, pulled using COMM ID 2821742
# https://rivers.codefornature.org/#/map
# pulled year 1983
# pulled wet season base flow for wet years
sj_wet_natural_flows_database_eff <- tibble(month = factor(month.abb[1:12], 
                                                        levels = month.abb[c(10, 11, 12, 1:9)]),
                                        flow_cfs = round(
                                          c(14566.04708, 19385.33385,
                                            28678.57203, 20474.8476, 
                                            16403.78026, 9007.285804, 
                                            6470.732515, 5570.717137, 
                                            4372.616735, 4358.370875, 
                                            6803.769081, 8721.041508), 0))

# Build Functional flow curve 
# still putting together what was done here
# first, fill in survival thresholds for migratory months
# then, set to 1000 (base flow for juvs? where is 1000 from?)
# then, optimize spawn wua. this also doesn't align with DSMhabitat::upper_mid_sac ?
# but I will use that for mine
# then, whatever is leftover allot to jan-april

# leftover from what? wet year?
# assuming the "allotment" is the total flow from the biop wet year
# sum(wet_year$flow_cfs)

# DSMhabitat::san_joaquin_river_instream |>
#   ggplot(aes(flow_cfs, y = FR_spawn_wua)) +
#   geom_line() +
#   geom_vline(aes(xintercept = 300), linetype = "dashed")

sj_functional_flows_for_model_wet <- tibble(month = factor(month.abb[1:12], 
                                                        levels = month.abb[c(10, 11, 12, 1:9)]),
                           # month_name = month.abb[month],
                           flow_cfs = c(45500, # Jan 
                                        45500, # March 
                                        45500, # March 
                                        45500, # April 
                                        1733, # May Spring recession flow (6,553 for mig surv ~50%)
                                        1733, # June Spring recession flow (6,553 for mig surv ~50%)
                                        1733, # July (6,553 for mig surv ~50%)
                                        2000, # Aug (50th percentile all years dry seasn baseflow from natural flows tool)
                                        2000, # Sept 
                                        300, # oct 
                                        300, # Nov 
                                        300 # Dec  
                                        ),
                           description = c("Utilize remaining water for FP habitat", # Try for fp rearing hab, less important - dependent on acre ft 
                                           "Utilize remaining water for FP habitat", # Optimize for  fp rearing habitat 
                                           "Utilize remaining water for FP habitat", # Optimize for  fp rearing habitat 
                                           "Utilize remaining water for FP habitat", # Optimize for fp rearing habitat 
                                           "Optimize for migratory survival (6,553 for mig surv ~50%)", # 
                                           "Optimize for migratory survival (6,553 for mig surv ~50%)", # Optimize for migratory survival (keep as high as possible but above mig surv thresholds)
                                           "Optimize for migratory survival (6,553 for mig surv ~50%)", # Optimize for migratory survival (keep as high as possible but above mig surv thresholds)
                                           "Set at 2000 to incrase acreage for avaliable for Juv flows", # set at 1000 to 
                                           "Set at 2000 to incrase acreage for avaliable for Juv flows",
                                           "Optimize for spawning habitat based on flow area WUA ", # Optimize for spawning habitat based on flow area WUA 
                                           "Optimize for spawning habitat based on flow area WUA ", # Optimize for spawning habitat 
                                           "Optimize for spawning habitat based on flow area WUA ")) # Optimize for spawning habitat 

# write_csv(functional_flows, "functional_flows_wet_year.csv")
knitr::kable(sj_functional_flows_for_model_wet)
```

The plot below shows the CalSIM modeled flows for the Sacramento River of an existing representative wet year: 1983. 1983 has the median water year index as provided by DWR out of all the "Wet" and "Above Normal" years within our model time frame.

```{r, echo = FALSE}
combined_data <- bind_rows((sj_functional_flows_for_model_wet |> mutate(Type = "Model EFF Wet Year")), 
                  (sj_wet_natural_flows_database_eff |> mutate(Type = "Natural Flows EFF")),
                  (wet_year |> mutate(Type = "CalSim")))
combined_data |> 
  ggplot(aes(x = month, y = flow_cfs, color = Type, group = Type)) + 
  geom_line() +
  theme_minimal() + 
  scale_color_manual(values = c("black", "lightblue", "gray")) +
  labs(x = "Month", 
       y = "Flow CFS", 
       title = "Wet Year - Model Constructed Salmonid Monthly Flows")
```

The table below shows the annual total volume of water needed for each flow scenario. The wet year EFF uses more water than the Natural Flows database EFF and then the representative CalSIM year. 

```{r, echo = FALSE}
compare_total_flow_volumne(wet_year, 
                           sj_wet_natural_flows_database_eff,
                           sj_functional_flows_for_model_wet)

```

#### Dry Year EFF

The table below shows monthly flow CFS for a dry year model EFF on the Sacramento River. This table was constructed using the Natural Flows database and model survival thresholds.

```{r, echo = FALSE}
# DRY YEAR EFF 
# Dry Years 
dry_year <- DSMflow::flows_cfs$biop_itp_2018_2019 |> 
  select(date, flow_cfs = `San Joaquin River`) |> 
  filter(year(date) == 1959) |>
    mutate(water_year = as.Date(ifelse(month(date) %in% 10:12, 
                             date,
                             as.Date(paste0(year(date) + 1, "-", month(date), "-", day(date))))),
         month = factor(month.abb[month(date)], levels = month.abb[c(10, 11, 12, 1:9)])) 

# dry_year |>
#   ggplot(aes(x = month, y = flow_cfs, group = 1)) +
#   geom_line()

# right now, pulled using COMM ID 2821742
# https://rivers.codefornature.org/#/map
# pulled year 1959
sj_dry_natural_flows_database_eff <- tibble(month = factor(month.abb[1:12], 
                                                        levels = month.abb[c(10, 11, 12, 1:9)]),
                                        flow_cfs = round(c(3731.569149, 7637.892379, 
                                                           5533.553387, 4005.806462, 
                                                           2616.398723, 2319.512281, 
                                                           2223.220098, 3122.946727, 
                                                           1153.635494, 1595.36601, 
                                                           1380.03132, 2005.553837)), 0)

# Build Functional flow curve 
# see notes on wet year for my method
# assuming the "allotment" is the total flow from the biop dry year
# sum(dry_year$flow_cfs)

sj_functional_flows_for_model_dry <- tibble(month = factor(month.abb[1:12], 
                                                        levels = month.abb[c(10, 11, 12, 1:9)]),
                           # month_name = month.abb[month],
                           flow_cfs = c(4000, # Jan Wet season base flow + Peak Magnitude flows, based on dry year monthly Natural Flows 
                                        4000, # Feb Wet season base flow + Peak Magnitude flows, based on dry year monthly Natural Flows 
                                        4000, # March Wet season base flow, based on dry year monthly Natural Flows 
                                        4000, # April Wet season base flow, based on dry year monthly Natural Flows 
                                        4339, # May Spring recession flow (keep migratory surv at ~20%)
                                        4339, # June Spring recession flow (keep migratory surv at ~20%)
                                        4339, # July half way between dry season baseflow and mig surv 20% threshold
                                        2000, # Aug - dry season baseflow 1 (keep migratory surv at ~20%), 
                                        300, # Sept - All fall runs juves have left system, use Natural Flows dry season base flow
                                        300, # oct - dry season baseflow, add some pulse flow in 
                                        300, # Nov add fall pulse flow, based on dry year monthly Natural Flows
                                        300 # Dec Begin dry season baseflow and medium flow, based on dry year monthly Natural Flows
                                         ),
                            description = c("Utilize remaining water for FP habitat", # Try for fp rearing hab, less important - dependent on acre ft 
                                           "Utilize remaining water for FP habitat", # Optimize for  fp rearing habitat 
                                           "Utilize remaining water for FP habitat", # Optimize for  fp rearing habitat 
                                           "Utilize remaining water for FP habitat", # Optimize for fp rearing habitat 
                                           "Keep high for migratory survival (3,069 for mig surv ~20%)", # 
                                           "Keep high for migratory survival (3,069 for mig surv ~20%)", # Optimize for migratory survival (keep as high as possible but above mig surv thresholds)
                                           "Keep high for migratory survival (3,069 for mig surv ~20%)", # Optimize for migratory survival (keep as high as possible but above mig surv thresholds)
                                           "Set summer base flow to 300 based on Natural flows tool and to incrase acreage for avaliable for Juv flows", # set at 1000 to 
                                           "Set summer base flow to 300 based on Natural flows tool and to incrase acreage for avaliable for Juv flows",
                                           "Optimize for spawning habitat based on flow area WUA ", # Optimize for spawning habitat based on flow area WUA 
                                           "Optimize for spawning habitat based on flow area WUA ", # Optimize for spawning habitat 
                                           "Optimize for spawning habitat based on flow area WUA ")) # Optimize for spawning habitat 

knitr::kable(sj_functional_flows_for_model_dry)

```

The plot below shows the CalSIM modeled flows for the San Joaquin River of an existing representative dry year: 1959. 1959 has the median water year index as provided by DWR out of all the "Critical", "Dry", and "Below Normal" years within our model time frame.


```{r, echo = FALSE}
combined_data_dry <- bind_rows((sj_functional_flows_for_model_dry |> mutate(Type = "Model EFF Wet Year")), 
                  (sj_dry_natural_flows_database_eff |> mutate(Type = "Natural Flows EFF")),
                  (dry_year |> mutate(Type = "CalSim")))
combined_data_dry |> 
  ggplot(aes(x = month, y = flow_cfs, color = Type, group = Type)) + 
  geom_line() +
  theme_minimal() + 
  scale_color_manual(values = c("black", "lightblue", "gray")) +
  labs(x = "Month", 
       y = "Flow CFS", 
       title = "Dry Year - Model Constructed Salmonid Monthly Flows")
```

The table below shows the annual total volume of water needed for each flow scenario. The dry year EFF uses less water than the Natural Flows database EFF and than the representative CalSIM year. 

```{r, echo = FALSE}
compare_total_flow_volumne(dry_year, 
                           sj_dry_natural_flows_database_eff,
                           sj_functional_flows_for_model_dry)
```

### Creating a synthetic flow time series

In order to incorporate these EFF into the model data inputs we combined wet and dry years to make a synthetic time series over the full 20 year period. We started by classifying years as "Dry" or "Wet" based on DWR water year type classifications. All "Dry", "Critical", or "Below Normal" years fall into the "Dry" category and all "Wet" or "Above Normal" years fall into the "Wet" category. 

The plot below shows this synthetic time series for the San Joaquin River over the 20 year simulation period.

```{r, echo = FALSE, warning=FALSE, message=FALSE}
# Wet dry sequence 
new_year_types <- water_year_types |> 
  mutate(year_type = ifelse(year_type %in% c("Dry", "Critical", "Below Normal"), 
                                             "Dry", "Wet")) |>
  select(year, year_type) 

dry <- sj_functional_flows_for_model_dry |> 
  mutate(year_type = "Dry",
         month = match(month.abb, month)) 

wet <- sj_functional_flows_for_model_wet |> 
  mutate(year_type = "Wet",
         month = match(month.abb, month)) 

wet_years <- 
  new_year_types |> 
  full_join(wet) 

dry_years <- new_year_types|> 
  full_join(dry) 

synthetic_eff_sj <- bind_rows(wet_years, dry_years) |> 
  mutate(date = as_date(paste0(year, "-", month, "-", ifelse(month == 2 & leap_year(year), 29,        lubridate::days_in_month(month))))) |> 
  # filter(year >= 1980, year <= 2000) |> 
  select(-year, -year_type, -month, -description, "Lower San Joaquin EFF" = flow_cfs, date) |>
  filter(!is.na(date)) 

low_sj_flows_eff_as_matrix <- synthetic_eff_sj |> 
    mutate(sjQcms = DSMflow::cfs_to_cms(`Lower San Joaquin EFF`),
          year = year(date),
          month = month(date)) |>
    filter(year >= 1980, year <= 2000) |> 
    arrange(date, ascending = TRUE) |> 
    select(-date, -`Lower San Joaquin EFF`) |>
    pivot_wider(names_from = year,
                values_from = sjQcms) |> 
    select(-month) |>
    as.matrix()
rownames(low_sj_flows_eff_as_matrix) <- month.abb[1:12]

usethis::use_data(synthetic_eff_sj, overwrite = TRUE)
synthetic_eff_sj |> 
  filter(year(date) >= 1979, year(date) <= 2000) |> 
  ggplot(aes(x = date, y = `Lower San Joaquin EFF`)) +
  geom_line(color = "blue4") + 
  theme_minimal() +
  labs(y = "Flow CFS", 
       x = "", 
       title = "Synthetic Model Constructed Salmonid Flows for San Joaquin River") 
```
```{r, echo = FALSE}
# Process data objects for use in modeling, data then used to create a scenario in transform_inputs_complete.R, source this file in transform_inputs_complete.R
```

```{r, echo = FALSE}
# compare to san joaquin calsim hydrograph, shape looks the same

# calsim_compare <- DSMflow::flows_cfs$biop_itp_2018_2019 |> 
#   select(date, flow_cfs = `San Joaquin River`) |>   
#   mutate(water_year = as.Date(ifelse(month(date) %in% 10:12, date,
#                                      as.Date(paste0(year(date) + 1, "-", month(date), "-", day(date))))),
#          month = factor(month.abb[month(date)], levels = month.abb[c(10, 11, 12, 1:9)])) |> 
#   mutate(water_year = ifelse(is.na(water_year), as.Date("1996-02-28"), water_year),
#          water_year = as.Date(water_year),
#          year = year(water_year)) |> 
#   left_join(new_year_types, by = c("year"))
# 
# calsim_compare |> 
#   filter(year >= 1979, year <= 2000) |> 
#   ggplot(aes(x = water_year, y = flow_cfs, color = year_type)) +
#   geom_line() + 
#   theme_minimal()
```

