---
title: "Functional Flow Analysis for the San Joaquin River"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Incorperating Effective Functional Flows into R2R

The Reorienting to Recovery Structured Decision Making group proposed a effective functional flow (EFF) scenario. To date, there is no available CalSIM EFF run so we cannot model full EFF on the entire system. This document outlines our approach to piloting an EFF on the San Joaquin River constructed to maximize salmonid survival.

This analysis is based on the analysis performed for the Sacramento River (`vignettes/EFF.Rmd`) and we do not replicate the methods here except to explain where they deviate for the San Joaquin.

### Review of Sacramento River EFF methods

A synthetic EFF hydrograph was developed for the Sacramento River using the following high-level steps: 

* Using the [Natural Flows Database](https://ceff.ucdavis.edu/sites/g/files/dgvnsk5566/files/media/documents/CEFF%20Technical%20Report%20Ver%201.0%20Mar_31_2021_DRAFT_FINAL%20for%20web.pdf), [California Environmental Flows Framework](https://ceff.ucdavis.edu/sites/g/files/dgvnsk5566/files/media/documents/CEFF%20Technical%20Report%20Ver%201.0%20Mar_31_2021_DRAFT_FINAL%20for%20web.pdf), and [Modeling Functional Flows in California's Rivers](https://www.frontiersin.org/articles/10.3389/fenvs.2022.787473/full#:~:text=Functional%20flow%20components%20in%20California,ecological%20processes%20in%20riverine%20ecosystems) as a basis, a monthly hydrograph was simulated to mimic  unimpaired historical flows

* Following this initial exercise, our synthetic hydrograph was further modified to maximize salmonid survival by accounting for flow-to-surivval thresholds for Chinook salmon as outlined in [Michel 2021](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.3498)

* Because this initial approach did not maximize *habitat* for salmon, the hydrograph was further modified to a) maximize spawning habitat from Oct-Dec, b) set a base flow for dry season (Aug-Sep) juvenile survival, c) optimize flows during the last three months of outmigration (May-July) using [Michel 2021](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.3498), and d) utilize additional flows for maximizing floodplain habitat from Jan-April.

### Adapting the Sacramento River EFF for the San Joaquin River

Our goals for developing an EFF hydrograph for the San Joaquin River were to maximize habitat and survival and reduce minimum flows when salmonids are not in the system.

In order to adapt the Sacramento EFF methodology, we need a flow-survival relationship for salmonids. Steps include: 

- scaling the existing flow-survival thresholds calculated by [Michel 2021](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.3498) for the Sacramento the San Joaquin
- modify the SJ survival model logic to be flow-based instead of temperature-based (so that the EFF has an effect)

The steps then become: 

- scale the Sacramento flow:survival relationship to produce flow-to-survival thresholds for salmonids in the San Joaquin
- use the Sacramento EFF logic with the SJ flow-to-survival thresholds 
- produce a synthetic EFF hydrograph for the SJ

### Methods

#### Scale flow-to-survival relationships

- plot SJ flow and Sacramento flow with Sacramento migratory thresholds
- "scale" them

```{r, echo = FALSE}
DSMflow::flows_cfs$biop_itp_2018_2019 |> 
  select(date, `Upper-mid Sacramento River`, `San Joaquin River`) |> 
  mutate(flow_relationship = `San Joaquin River` / `Upper-mid Sacramento River`) |> pivot_longer(`Upper-mid Sacramento River`:`San Joaquin River`,
               names_to = "watershed", values_to = "flow_cfs") |> 
  ggplot(aes(x = date, y = flow_cfs, color = watershed)) + 
  geom_line() +
  theme_minimal()

scale_thresholds <- DSMflow::flows_cfs$biop_itp_2018_2019 |> 
  select(date, `Upper-mid Sacramento River`, `San Joaquin River`) |> 
  mutate(flow_relationship = `San Joaquin River` / `Upper-mid Sacramento River`) |> pivot_longer(`Upper-mid Sacramento River`:`San Joaquin River`,
               names_to = "watershed", values_to = "flow_cfs") |> 
  filter(watershed == "San Joaquin River") |> 
  glimpse()

scale_thresholds |> 
  ggplot(aes(x = date, y = flow_relationship)) + 
  geom_line() +
  theme_minimal() 
# ranges from 0:1, 3 in some later years and model years

# filter to model years and then average to get scaling factor
san_joaquin_scaling_factor <- scale_thresholds |> 
  filter(year(date) %in% 1980:2000) |> 
  summarise(scaling_factor = mean(flow_relationship)) |> 
  pull(scaling_factor)

# now calculate flow required for survival using the scaling factor to
# adjust flows
flow_to_survival_thresholds <- tibble("Survival" = c(3, 18.9, 50.8),
                                      "Sacramento Flow Threshold" = c(4279, 10712, 22872)) |> 
  mutate(`San Joaquin Flow Threshold` = round(`Sacramento Flow Threshold` * san_joaquin_scaling_factor, 0))

knitr::kable(flow_to_survival_thresholds)
```

```{r, echo=FALSE}
# Helper / Set Up Code
# Compare total flow volumne function 
compare_total_flow_volumne <- function(calsim_flows, natural_flow_flows, proposed_eff){
  all_flows <- bind_rows(calsim_flows |> mutate(type = "calsim"), 
                         natural_flow_flows |> mutate(type = "Natural flows Database flows"),
                         proposed_eff |> mutate(type = "Proposed Model CSF")) |> 
    select(month, flow_cfs, type) |> 
    mutate(acre_ft_per_month = flow_cfs * 60.3707) |> #TODO confirm conversion 
    group_by(type) |> 
    summarise(annual_acre_feet = sum(acre_ft_per_month)) |> 
    rename(Scenario = type, "Annual Acre Ft Water" = annual_acre_feet) 
  
  return(knitr::kable(all_flows))
}

# Find Representative water year types
water_year_types <- waterYearType::water_year_indices |> 
  filter(location == "Sacramento Valley", WY %in% c(1922:2003)) |> 
  select(year = WY, index = Index, year_type = Yr_type) 


median_wet_year_index <- water_year_types |> 
  filter(year_type %in% c("Wet", "Above Normal")) |> 
  pull(index) |> median()

represenative_wet_years <- water_year_types |>
  mutate(distance_from_median = abs(index - median_wet_year_index)) |> 
  filter(distance_from_median < 1) 
  
# Representative Wet years are 1984 & 1996 - both .13 away 


median_dry_year_index <- water_year_types |> 
  filter(year_type %in% c("Critical", "Dry", "Below Normal")) |> 
  pull(index) |> median()

represenative_dry_years <- water_year_types |>
  mutate(distance_from_median = abs(index - median_dry_year_index)) |> 
  filter(distance_from_median < 1) 

# Representative Wet years are 1987 & 1994 - both .42 away

```

### Proposed Model Constructed Salmonid Flows (CSF) {.tabset}

#### Wet Year CSF

The table below shows monthly flow CFS for a wet year model CSF on the Sacramento River. This table was constructed using the Natural Flows database and model survival thresholds.

```{r, echo = FALSE}
# WET YEAR EFF 
# NATURAL FLOWS TOOL COMID: 2851441

# For calsim flow picked wet year, wettest year on reccord 
# TODO find average wet year 
wet_year <- DSMflow::flows_cfs$biop_itp_2018_2019 |> 
  select(date, flow_cfs = `Upper-mid Sacramento River`) |> 
  filter(year(date) == 1996) |>
  mutate(water_year = as.Date(ifelse(month(date) %in% 10:12, 
                             date,
                             as.Date(paste0(year(date) + 1, "-", month(date), "-", day(date))))),
         month = factor(month.abb[month(date)], levels = month.abb[c(10, 11, 12, 1:9)])) 

# wet_year |> 
#   ggplot(aes(x = water_year, y = flow_cfs)) + 
#   geom_line()

```

```{r,echo = FALSE}
sac_wet_natural_flows_database_eff <- tibble(month = factor(month.abb[1:12], 
                                                        levels = month.abb[c(10, 11, 12, 1:9)]),
                                        flow_cfs = c(62000, 48900, 43200,
                                                     37300, 19200, 9270,
                                                     3140, 1680, 1280, 
                                                     3760, 11600, 30000))

# Build Functional flow curve 
sac_functional_flows_for_model_wet <- tibble(month = factor(month.abb[1:12], 
                                                        levels = month.abb[c(10, 11, 12, 1:9)]),
                           # month_name = month.abb[month],
                           flow_cfs = c(40000, # Jan 
                                        40000, # March 
                                        40000, # March 
                                        40000, # April 
                                        10700, # May Spring recession flow (10,700 for mig surv ~50%)
                                        10700, # June Spring recession flow (10,700 for mig surv ~50%)
                                        10700, # July (10,700 for mig surv ~50%)
                                        1000, # Aug 
                                        1000, # Sept 
                                        4750, # oct 
                                        4750, # Nov 
                                        4750 # Dec  
                                         ),
                           description = c("Utilize remaining water for FP habitat", # Try for fp rearing hab, less important - dependent on acre ft 
                                           "Utilize remaining water for FP habitat", # Optimize for  fp rearing habitat 
                                           "Utilize remaining water for FP habitat", # Optimize for  fp rearing habitat 
                                           "Utilize remaining water for FP habitat", # Optimize for fp rearing habitat 
                                           "Optimize for migratory survival (10,700 for mig surv ~50%)", # 
                                           "Optimize for migratory survival (10,700 for mig surv ~50%)", # Optimize for migratory survival (keep as high as possible but above mig surv thresholds)
                                           "Optimize for migratory survival (10,700 for mig surv ~50%)", # Optimize for migratory survival (keep as high as possible but above mig surv thresholds)
                                           "Set at 1000 to incrase acreage for avaliable for Juv flows", # set at 1000 to 
                                           "Set at 1000 to incrase acreage for avaliable for Juv flows",
                                           "Optimize for spawning habitat based on flow area WUA ", # Optimize for spawning habitat based on flow area WUA 
                                           "Optimize for spawning habitat based on flow area WUA ", # Optimize for spawning habitat 
                                           "Optimize for spawning habitat based on flow area WUA ")) # Optimize for spawning habitat 

# write_csv(functional_flows, "functional_flows_wet_year.csv")
knitr::kable(sac_functional_flows_for_model_wet)
```

The plot below shows the CalSIM modeled flows for the Sacramento River of an existing representative wet year: 1996. 1996 has the median water year index as provided by DWR out of all the "Wet" and "Above Normal" years within our model time frame.

```{r, echo = FALSE}
combined_data <- bind_rows((sac_functional_flows_for_model_wet |> mutate(Type = "Model CSF Wet Year")), 
                  (sac_wet_natural_flows_database_eff |> mutate(Type = "Natural Flows EFF")),
                  (wet_year |> mutate(Type = "CalSim")))
combined_data |> 
  ggplot(aes(x = month, y = flow_cfs, color = Type, group = Type)) + 
  geom_line() +
  theme_minimal() + 
  scale_color_manual(values = c("black", "lightblue", "gray")) +
  labs(x = "Month", 
       y = "Flow CFS", 
       title = "Wet Year - Model Constructed Salmonid Monthly Flows")
```

The table below shows the annual total volume of water needed for each flow scenario. The wet year CSF uses more water than the Natural Flows database EFF and then the representative CalSIM year. 

```{r, echo = FALSE}
compare_total_flow_volumne(wet_year, 
                           sac_wet_natural_flows_database_eff,
                           sac_functional_flows_for_model_wet)

```

#### Dry Year CSF

The table below shows monthly flow CFS for a dry year model CSF on the Sacramento River. This table was constructed using the Natural Flows database and model survival thresholds.

```{r, echo = FALSE}
# DRY YEAR EFF 
# Dry Years 
dry_year <- DSMflow::flows_cfs$biop_itp_2018_2019 |> 
  select(date, flow_cfs = `Upper-mid Sacramento River`) |> 
  filter(year(date) == 1987) |>
    mutate(water_year = as.Date(ifelse(month(date) %in% 10:12, 
                             date,
                             as.Date(paste0(year(date) + 1, "-", month(date), "-", day(date))))),
         month = factor(month.abb[month(date)], levels = month.abb[c(10, 11, 12, 1:9)])) 

# dry_year |> 
#   ggplot(aes(x = month, y = flow_cfs, group = 1)) + 
#   geom_line()

sac_dry_natural_flows_database_eff <- tibble(month = factor(month.abb[1:12], 
                                                        levels = month.abb[c(10, 11, 12, 1:9)]),
                                        flow_cfs = c(14100, 15700, 15600, 
                                                     9560, 4870, 2100, 
                                                     781, 626, 564, 
                                                     2230, 6060, 10100))

# Build Functional flow curve 
sac_functional_flows_for_model_dry <- tibble(month = factor(month.abb[1:12], 
                                                        levels = month.abb[c(10, 11, 12, 1:9)]),
                           # month_name = month.abb[month],
                           flow_cfs = c(18500, # Jan Wet season base flow + Peak Magnitude flows, based on dry year monthly Natural Flows 
                                        18500, # Feb Wet season base flow + Peak Magnitude flows, based on dry year monthly Natural Flows 
                                        18500, # March Wet season base flow, based on dry year monthly Natural Flows 
                                        18500, # April Wet season base flow, based on dry year monthly Natural Flows 
                                        4500, # May Spring recession flow (keep migratory surv at ~20%)
                                        4500, # June Spring recession flow (keep migratory surv at ~20%)
                                        4500, # July half way between dry season baseflow and mig surv 20% threshold
                                        500, # Aug - dry season baseflow 1 (keep migratory surv at ~20%), 
                                        500, # Sept - All fall runs juves have left system, use Natural Flows dry season base flow
                                        4750, # oct - dry season baseflow, add some pulse flow in 
                                        4750, # Nov add fall pulse flow, based on dry year monthly Natural Flows
                                        4750 # Dec Begin dry season baseflow and medium flow, based on dry year monthly Natural Flows
                                         ),
                            description = c("Utilize remaining water for FP habitat", # Try for fp rearing hab, less important - dependent on acre ft 
                                           "Utilize remaining water for FP habitat", # Optimize for  fp rearing habitat 
                                           "Utilize remaining water for FP habitat", # Optimize for  fp rearing habitat 
                                           "Utilize remaining water for FP habitat", # Optimize for fp rearing habitat 
                                           "Keep high for migratory survival (4,350 for mig surv ~20%)", # 
                                           "Keep high for migratory survival (4,350 for mig surv ~20%)", # Optimize for migratory survival (keep as high as possible but above mig surv thresholds)
                                           "Keep high for migratory survival (4,350 for mig surv ~20%)", # Optimize for migratory survival (keep as high as possible but above mig surv thresholds)
                                           "Set summer base flow to 550 based on Natural flows tool and to incrase acreage for avaliable for Juv flows", # set at 1000 to 
                                           "Set summer base flow to 550 based on Natural flows tool and to incrase acreage for avaliable for Juv flows",
                                           "Optimize for spawning habitat based on flow area WUA ", # Optimize for spawning habitat based on flow area WUA 
                                           "Optimize for spawning habitat based on flow area WUA ", # Optimize for spawning habitat 
                                           "Optimize for spawning habitat based on flow area WUA ")) # Optimize for spawning habitat 

knitr::kable(sac_functional_flows_for_model_dry)

```

The plot below shows the CalSIM modeled flows for the Sacramento River of an existing representative dry year: 1987. 1987 has the median water year index as provided by DWR out of all the "Critical", "Dry", and "Below Normal" years within our model time frame.


```{r, echo = FALSE}
combined_data_dry <- bind_rows((sac_functional_flows_for_model_dry |> mutate(Type = "Model CSF Wet Year")), 
                  (sac_dry_natural_flows_database_eff |> mutate(Type = "Natural Flows EFF")),
                  (dry_year |> mutate(Type = "CalSim")))
combined_data_dry |> 
  ggplot(aes(x = month, y = flow_cfs, color = Type, group = Type)) + 
  geom_line() +
  theme_minimal() + 
  scale_color_manual(values = c("black", "lightblue", "gray")) +
  labs(x = "Month", 
       y = "Flow CFS", 
       title = "Dry Year - Model Constructed Salmonid Monthly Flows")
```

The table below shows the annual total volume of water needed for each flow scenario. The dry year CSF uses less water than the Natural Flows database EFF and than the representative CalSIM year. 

```{r, echo = FALSE}
compare_total_flow_volumne(dry_year, 
                           sac_dry_natural_flows_database_eff,
                           sac_functional_flows_for_model_dry)

```

### Creating a synthetic flow timeserries

In order to incorporate these CSF into the model data inputs we combined wet and dry years to make a synthetic time series over the full 20 year period. We started by classifying years as "Dry" or "Wet" based on DWR water year type classifications. All "Dry", "Critical", or "Below Normal" years fall into the "Dry" category and all "Wet" or "Above Normal" years fall into the "Wet" category. 

The plot below shows this synthetic time series for the Sacramento River over the 20 year simulation period.

```{r, echo = FALSE, warning=FALSE, message=FALSE}
# Wet dry sequence 
new_year_types <- water_year_types |> 
  mutate(year_type = ifelse(year_type %in% c("Dry", "Critical", "Below Normal"), 
                                             "Dry", "Wet")) |>
  select(year, year_type) 

dry <- sac_functional_flows_for_model_dry |> 
  mutate(year_type = "Dry",
         month = match(month.abb, month)) 

wet <- sac_functional_flows_for_model_wet |> 
  mutate(year_type = "Wet",
         month = match(month.abb, month)) 

wet_years <- 
  new_year_types |> 
  full_join(wet) 

dry_years <- new_year_types|> 
  full_join(dry) 

synthetic_eff_sac <- bind_rows(wet_years, dry_years) |> 
  mutate(date = as_date(paste0(year, "-", month, "-", ifelse(month == 2 & leap_year(year), 29,        lubridate::days_in_month(month))))) |> 
  # filter(year >= 1980, year <= 2000) |> 
  select(-year, -year_type, -month, -description, "Upper Sacramento River EFF" = flow_cfs, date) |>
  filter(!is.na(date)) 

up_sac_flows_eff_as_matrix <- synthetic_eff_sac |> 
    mutate(upsacQcms = DSMflow::cfs_to_cms(`Upper Sacramento River EFF`),
          year = year(date),
          month = month(date)) |>
    filter(year >= 1980, year <= 2000) |> 
    arrange(date, ascending = TRUE) |> 
    select(-date, -`Upper Sacramento River EFF`) |>
    pivot_wider(names_from = year,
                values_from = upsacQcms) |> 
    select(-month) |>
    as.matrix()
rownames(up_sac_flows_eff_as_matrix) <- month.abb[1:12]

usethis::use_data(synthetic_eff_sac, overwrite = TRUE)
synthetic_eff_sac |> 
  filter(year(date) >= 1979, year(date) <= 2000) |> 
  ggplot(aes(x = date, y = `Upper Sacramento River EFF`)) +
  geom_line(color = "blue4") + 
  theme_minimal() +
  labs(y = "Flow CFS", 
       x = "", 
       title = "Synthetic Model Constructed Salmonid Flows for Sacramento River") 
```
```{r, echo = FALSE}
# Process data objects for use in modeling, data then used to create a scenario in transform_inputs_complete.R, source this file in transform_inputs_complete.R
```

